---
title: "Introduction to R"
subtitle: Hands-on adapted from [Introduction to R and Bioconductor](https://carpentries-incubator.github.io/bioc-intro/index.html)
author: Hosna Baniadam and Nick Hirschmüller 
output:
  BiocStyle::html_document:
      toc: yes
      toc_depth: 2
      toc_float: yes
      df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE, cache = FALSE)
```

# What is this part of the hands-on about?

**Questions:**

- First commands in R

**Objectives:**

- Define the following terms as they relate to R: object, assign, call, function, arguments, options.
- Assign values to objects in R.
- Learn how to _name_ objects
- Use comments to inform script.
- Solve simple arithmetic operations in R.
- Call functions and use arguments to change their default options.
- Inspect the content of vectors and manipulate their content.
- Subset and extract values from vectors.
- Analyze vectors with missing data.

**Keypoints:**

- How to interact with R

# Creating objects in R

You can get output from R simply by typing math in the console:

```{r}
3 + 5
12 / 7
```

However, to do useful and interesting things, we need to assign _values_ to
_objects_. To create an object, we need to give it a name followed by the
assignment operator `<-`, and the value we want to give it:

```{r}
weight_kg <- 55
```

`<-` is the assignment operator. It assigns values on the right to
objects on the left. So, after executing `x <- 3`, the value of `x` is
`3`. The arrow can be read as 3 **goes into** `x`.  For historical
reasons, you can also use `=` for assignments, but it is good practice to always 
use `<-` for assignments.

## Naming variables

There are some restrictions on the name that you give an object:
They cannot start with a number (`2x` is not valid, but `x2`
is) and some names cannot be used because they
are the names of fundamental functions in R (e.g., `if`, `else`, `for`, ...). 
Furthermore, its best to not use other function names. Ideally, you want
your object names to be explicit and not too long. It's important to be consistent 
in the styling of your code (where you put spaces, how you name objects, etc.). 
Using a consistent coding style makes your code clearer to read for your
future self and your collaborators. In R, some popular style guides
are [Google's](https://google.github.io/styleguide/Rguide.xml), the
[tidyverse's](http://style.tidyverse.org/) style and the [Bioconductor
style
guide](https://bioconductor.org/developers/how-to/coding-style/). The
tidyverse's is very comprehensive and may seem overwhelming at
first. You can install the
[**`lintr`**](https://github.com/jimhester/lintr) package to
automatically check for issues in the styling of your code.

**Objects vs. variables**: What are known as `objects` in `R` are
known as `variables` in many other programming languages. Depending
on the context, `object` and `variable` can have drastically
different meanings. However, in this lesson, the two words are used
synonymously. For more information
[see here.](https://cran.r-project.org/doc/manuals/r-release/R-lang.html#Objects)


When assigning a value to an object, R does not print anything. You
can force R to print the value by using parentheses or by typing the
object name:
```{r}
weight_kg <- 55    # doesn't print anything
(weight_kg <- 55)  # but putting parenthesis around the call prints the value of `weight_kg`
weight_kg          # and so does simply typing the name of the object
```

Now that R has `weight_kg` in memory, we can do arithmetic with it. For
instance, we may want to convert this weight into pounds (weight in pounds is 2.2 times the weight in kg):

```{r}
2.2 * weight_kg
```

We can also change an object's value by assigning it a new one:

```{r}
weight_kg <- 57.5
2.2 * weight_kg
```

This means that assigning a value to one object does not change the values of
other objects  For example, let's store the animal's weight in pounds in a new
object, `weight_lb`:

```{r}
weight_lb <- 2.2 * weight_kg
```

and then change `weight_kg` to 100.

```{r}
weight_kg <- 100
```

- **Challenge:**

What do you think is the current content of the object `weight_lb`?
126.5 or 220?

- **Challenge**

What are the values after each statement in the following?

```{r, eval = FALSE}
mass <- 47.5            # mass?
age  <- 122             # age?
mass <- mass * 2.0      # mass?
age  <- age - 20        # age?
mass_index <- mass/age  # mass_index?
```

# Comments

The comment character in R is `#`, anything to the right of a `#` in a
script will be ignored by R. It is useful to leave notes, and
explanations in your scripts.

RStudio makes it easy to comment or uncomment a paragraph: after
selecting the lines you want to comment, press at the same time on
your keyboard <kbd>Ctrl</kbd> + <kbd>Shift</kbd> + <kbd>C</kbd>. If
you only want to comment out one line, you can put the cursor at any
location of that line (i.e. no need to select the whole line), then
press <kbd>Ctrl</kbd> + <kbd>Shift</kbd> + <kbd>C</kbd>.



# Functions and their arguments

Functions are "canned scripts" that automate more complicated sets of commands
including operations assignments, etc. Many functions are predefined, or can be
made available by importing R *packages* (more on that later). A function
usually gets one or more inputs called *arguments*. Functions often (but not
always) return a *value*. A typical example would be the function `sqrt()`. The
input (the argument) must be a number, and the return value (in fact, the
output) is the square root of that number. Executing a function ('running it')
is called *calling* the function. An example of a function call is:

```{r, eval=FALSE}
a <- 9
b <- sqrt(a)
```
Arguments can be anything, not only numbers, but also other
objects. Exactly what each argument means differs per function, and must be
looked up in the documentation (see below). Some functions take arguments which
may either be specified by the user, or, if left out, take on a *default* value.

Let's try a function that can take multiple arguments: `round()`.

```{r, results='show'}
round(3.14159)
```

Here, we've called `round()` with just one argument, `3.14159`, and it has
returned the value `3`.  That's because the default is to round to the nearest
whole number. If we want more digits we can see how to do that by getting
information about the `round` function.  We can use `?round` to get more information
and associated arguments of the function. 

```{r, eval=FALSE}
?round
```

We see that if we want a different number of digits, we can
type `digits=2` or however many we want.

```{r, results='show'}
round(3.14159, digits = 2)
```

If you provide the arguments in the exact same order as they are defined by the function, you
don't have to name them:

```{r, results='show'}
round(3.14159, 2)
```

It's good practice to put the non-optional arguments (like the number you're
rounding) first in your function call, and to specify the names of all optional
arguments.  If you don't, someone reading your code might have to look up the
definition of a function with unfamiliar arguments to understand what you're
doing. By specifying the name of the arguments you are also safeguarding
against possible future changes in the function interface, which may
potentially add new arguments in between the existing ones.

# Vectors and data types

A vector is the most common and basic data type in R, and is pretty much
the workhorse of R. A vector is composed by a series of values, such as
numbers or characters. We can assign a series of values to a vector using
the `c()` function. For example we can create a vector of animal weights and assign
it to a new object `weight_g`:

```{r}
weight_g <- c(50, 60, 65, 82)
weight_g
```

A vector can also contain characters:

```{r}
molecules <- c("dna", "rna", "protein")
molecules
```

The quotes around "dna", "rna", etc. are essential here. Without the
quotes R will assume there are objects called `dna`, `rna` and
`protein`. As these objects don't exist in R's memory, there will be
an error message.

There are many functions that allow you to inspect the content of a
vector. `length()` tells you how many elements are in a particular vector:

```{r}
length(weight_g)
length(molecules)
```

An important feature of a vector, is that all of the elements are the
same type of data, i.e. all have to be either *strings*, *integers* (whole numbers)
or *doubles* (real numbers). 
The function `str()` provides an overview of the structure of an
object and its elements. It is a useful function when working with
large and complex objects:

```{r}
str(weight_g)
str(molecules)
```
You can use the `c()` function to add other elements to your vector:

```{r}
weight_g <- c(weight_g, 90) # add to the end of the vector
weight_g <- c(30, weight_g) # add to the beginning of the vector
weight_g
```

In the first line, we take the original vector `weight_g`, add the
value `90` to the end of it, and save the result back into
`weight_g`. Then we add the value `30` to the beginning, again saving
the result back into `weight_g`.

Vectors are one of the many **data structures** that R uses. Other
important ones are lists (`list`), matrices (`matrix`), data frames
(`data.frame`), factors (`factor`) and arrays (`array`).

# Subsetting vectors

If we want to extract one or several values from a vector, we must
provide one or several indices in square brackets. For instance:

```{r, results='show'}
molecules <- c("dna", "rna", "peptide", "protein")
molecules[2]
molecules[c(3, 2)]
```

R indices start at 1. Programming languages like Fortran, MATLAB,
Julia, and R start counting at 1, because that's what human beings
typically do. Languages in the C family (including C++, Java, Perl,
and Python) count from 0 because that's simpler for computers to do.

Finally, it is also possible to get all the elements of a vector
except some specified elements using negative indices:

```{r}
molecules # all molecules
molecules[-1] # all but the first one
molecules[-c(1, 3)] # all but 1st/3rd ones
molecules[c(-1, -3)] # all but 1st/3rd ones
```

# Conditional subsetting

Another common way of subsetting is by using a logical vector. 
A logical vector, is a vector that contains only `TRUE` or `FALSE` as their elements.
`TRUE` will select the element with the same index, while `FALSE` will not:

```{r, }
weight_g <- c(21, 34, 39, 54, 55)
weight_g[c(TRUE, FALSE, TRUE, TRUE, FALSE)]
```

Typically, these logical vectors are not typed by hand, but are the
output of other functions or logical tests. For instance, if you
wanted to select only the values above 50:

```{r, }
# will return logicals with TRUE for the indices that meet
# the condition
weight_g > 50
# so we can use this to select only the values above 50
weight_g[weight_g > 50]
```

You can combine multiple tests using `&` (both conditions are true,
AND) or `|` (at least one of the conditions is true, OR):

```{r, results='show'}
weight_g[weight_g < 30 | weight_g > 50]
```
A common task is to search for certain strings in a vector.  One could
use the "or" operator `|` to test for equality to multiple values, but
this can quickly become tedious. The function `%in%` allows you to
test if any of the elements of a search vector are found:

```{r, }
molecules <- c("dna", "rna", "protein", "peptide")
molecules[molecules == "rna" | molecules == "dna"] # returns both rna and dna
molecules %in% c("rna", "dna", "metabolite", "peptide", "glycerol")
molecules[molecules %in% c("rna", "dna", "metabolite", "peptide", "glycerol")]
```
# Names

It is possible to name each element of a vector. The code chunk below
shows an initial vector without any names, how names are set, and
retrieved.

```{r}
x <- c(1, 5, 3, 5, 10)
names(x) # no names
names(x) <- c("A", "B", "C", "D", "E")
names(x) # now we have names
x
```

When a vector has names, it is possible to access elements by their
name, in addition to their index.

```{r}
x[c(1, 3)]
x[c("A", "C")]
```

# Missing data

As R was designed to analyze datasets, it includes the concept of
missing data. Missing data are represented in vectors as `NA`.

When doing operations on numbers, most functions will return `NA` if
the data you are working with include missing values. This feature
makes it harder to overlook the cases where you are dealing with
missing data.  You can add the argument `na.rm = TRUE` to calculate
the result while ignoring the missing values.

```{r}
heights <- c(2, 4, 4, NA, 6)
mean(heights)
max(heights)
mean(heights, na.rm = TRUE)
max(heights, na.rm = TRUE)
```

If your data include missing values, you may want to become familiar
with the functions `is.na()` and `complete.cases()`. See
below for examples.

```{r}
# Extract those elements which are not missing values.
is.na(heights)
heights[!is.na(heights)] # the "!" inverts bool vectors, i.e. TRUE becomes FALSE and vice versa

# Extract those elements which are complete cases.
heights[complete.cases(heights)]
```

- **Challenge**

1. Using this vector of heights in inches, create a new vector with the NAs removed.

```{r}
heights <- c(63, 69, 60, 65, NA, 68, 61, 70, 61, 59, 64, 69, 63, 63, NA, 72, 65, 64, 70, 63, 65)
```

2. Use the function `median()` to calculate the median of the `heights` vector.
3. Use R to figure out how many people in the set are taller than 67 inches.

# Generating vectors {#sec:genvec}
## Replicate elements

The `rep` function allows to repeat a value a certain number of
times. If we want to initiate a vector of numerics of length 5 with
the value -1, for example, we could do the following:

```{r}
rep(-1, 5)
```

Similarly, to generate a vector populated with missing values, which
is often a good way to start, without setting assumptions on the data
to be collected:

```{r}
rep(NA, 5)
```

`rep` can take vectors of any length as input (above, we used vectors
of length 1) and any type. For example, if we want to repeat the
values 1, 2 and 3 five times, we would do the following:

```{r}
rep(c(1, 2, 3), 5)
```
- **Challenge**

What if we wanted to repeat the values 1, 2 and 3 five times, but
obtain five 1s, five 2s and five 3s in that order? There are two
possibilities - see `?rep` or `?sort` for help.

## Sequence generation

Another very useful function is `seq`, to generate a sequence of
numbers. For example, to generate a sequence of integers from 1 to 20
by steps of 2, one would use:

```{r}
seq(from = 1, to = 20, by = 2)
```

The default value of `by` is 1 and, given that the generation of a
sequence of one value to another with steps of 1 is frequently used,
there's a shortcut:

```{r}
seq(1, 5, 1)
seq(1, 5) # default by
1:5
```

To generate a sequence of numbers from 1 to 20 of final length of 3,
one would use:

```{r}
seq(from = 1, to = 20, length.out = 3)
```

## Random samples and permutations

A last group of useful functions are those that generate random
data. The first one, `sample`, generates a random permutation of
another vector. For example, to draw a random order to 10 students
oral exam, I first assign each student a number from 1 to ten (for
instance based on the alphabetic order of their name) and then:

```{r}
sample(1:10)
```

Without further arguments, `sample` will return a permutation of all
elements of the vector. If I want a random sample of a certain size, I
would set this value as the second argument. Below, I sample 5 random
letters from the alphabet contained in the pre-defined `letters` vector:

```{r}
letters
sample(letters, 5)
```

If I wanted an output larger than the input vector, or being able to
draw some elements multiple times, I would need to set the `replace`
argument to `TRUE`:

```{r}
sample(1:5, 10, replace = TRUE)
```

# Outlook
Now that we have learned how to write scripts, and the basics of R's
data structures, we are ready to start working with larger data, and
learn about data frames.
